Typing a linear \picalc

### 2020-03-29 paper/paper.tex
Small improvements
Use tuples for io multiplicities

% Fields
\newcommand{\op}[3]{#1 \; \field{\coloneqq} \; #2 \; \field{\cdot} \; #3}
\newcommand{\li}{\field{\ell_i}}
\newcommand{\lo}{\field{\ell_o}}
\newcommand{\lz}{\field{\ell_{\o}}}
\newcommand{\lio}{\field{\ell_{\#}}}
\newcommand{\Balanced}{\field{Balanced}}

@@

In its usual representation, the linear \picalc{} distinguishes between \emph{capabilities} -- $\ell_{\o}$ for cannot send nor receive, $\ell_i$ for input, $\ell_o$ for output, and $\ell_\#$ for both input and output -- and multiplicities -- $0$ or $1$.
This introduces some unnecessary semantic duplication: a channel with multiplicity 1 and capability $\cdot \ell_{\o}$ is indistinguishable, in usage, from a channel with multiplicity 0.
In this paper, we merge multiplicities and capabilities into a single type and define the algebraic rules for a generalized resource-aware type system on the \picalc{}.
As such, it can be used to model shared, linear, or gradual types.

The algebra has a single binary operator $(\cdot)$ that crushes together capabilities, and must satisfy the following characteristics:
\begin{description}

\item [Partiality]
  Not all capabilities are necessarily compatible: in the linear \picalc{} the capability $\ell_i$ cannot be split into $\ell_i$ and $\ell_i$.
\item [Functionality]
  Given two capabilities $y$ and $z$, it must be computable whether a third capability $x$ exists such that $\op{x}{y}{z}$.
  If such a $x$ exists, it must be unique.

\item [Injectivity]
  To prove framing (\S \ref{framing}) we need to be able to compute the \emph{subtraction} of capabilities.
  That is, $x - z \equiv y$ where $y$ is uniquely determined iff $\op{x}{y}{z}$ where $y$ is uniquely determined.

@@

\item [Distinguished elements]
  There must be a neutral element $\ell_{\o}$ absorbed on both sides.
  There need to be two special elements $\ell_i$ and $\ell_o$ that will be subtracted every time input and output (respectively) happens.
  These two elements need to be compatible, and we will refer to the result as $\ell_\#$.
\end{description}

In Figure \ref{capabilities} we present an algebra $\Quantifier_C$ on the carrier set $C$ in the form of a terniary relation, satisfying all the properties mentioned above.

\todo{Ornela: in linear pi if you write $\li$ the $\ell$ already means that the multiplicity is 1, meaning linear. 
We need to distinguish between \emph{capabilities} i, o or $\sharp$ and \emph{multiplicities} 0, 1 or $\omega$. So, the $\ell$ cannot be $\omega$, hence shared. You should use only i, o or $\sharp$ in Fig 7. and use $\ell$ as a syntactic sugar for i, o or $\sharp$ with multiplicity 1.}

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\lz                    &:{} &                      &        & C \\
  &\li                    &:{} &                      &        & C \\
  &\lo                    &:{} &                      &        & C \\
  &\lio                   &:{} &                      &        & C \\
  &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
  &\field{\cdot-join}     &:{} &                      &        & \op{\lio}{\li}{\lo} \\
  &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
  &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  &\field{\cdot-id^l}     &:{} &\forall x             & \to \; & \op{x}{\lz}{x} \\
  &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
  &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
\end{aligned}
\end{equation}
\caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid with special elements $\ell_i$, $\ell_o$ and $\ell_\#$.}
\label{capabilities}
\end{figure}

To allow for several capability algebras to coexist on the same type system, in Figure \ref{indexed-capabilities} we index the algebra carriers by an index $\Idx$.
Such an index is a type with one element for each algebra of our type system.
To be able to prove subject congruence, we require for at least one such algebra to exist.
Finally, the field $\field{QUANTIFIERS}$ requires for each index $i$ to have an algebra defined on the carrier set $\constr{\Carrier_i}$.

\begin{figure}[h]
\begin{equation}
\begin{split}
  &\field{IDX}          &:{} &\Set \\
  &\field{\exists IDX}  &:{} &\field{IDX} \\
  &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
  &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
\end{split}
\end{equation}
\caption{Indexed set of partial commutative monoids}
\label{indexed-capabilities}
\end{figure}

@@

@@

The carrier $C$ is implemented as a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
All of $\lz$, $\li$, $\lo$ and $\lio$ are instantiated to $\constr{\omega}$.
The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
All other laws are trivially satisfied.

@@

The carrier $C$ is implemented as a type $\type{Four}$ with the trivial elements $\constr{\ell_{\o}}$, $constr{\ell_i}$, $\constr{\ell_o}$ and $\constr{\ell_\#}$.
The monoidal operation has the element $\constr{\ell_{\o}}$ as neutral on both sides, and the element $\constr{\ell_\#}$ splitting into $\constr{\ell_i}$ and $\constr{\ell_o}$ (or $\constr{\ell_o}$ and $\constr{\ell_i}$ -- and is uninhabited in every other case.
All other rules follow trivially.

@@

The carrier $C$ is implemented as a tuple $\N \type{\times} \N$ of natural numbers.
The element $\lz$ corresponds to the tuple $(\constr{0} \constr{,} \constr{0})$, $\li$ to $(\constr{1} \constr{,} \constr{0})$, $\lo$ to $(\constr{0} \constr{,} \constr{1})$, and $\lio$ to $(\constr{1} \constr{,} \constr{1})$.
The partial monoid is defined exactly when $\op{(x_l + y_l \constr{,} x_r + y_r)}{(x_l \constr{,} x_r)}{(y_l \constr{,} y_r)}$.
All other rules follow trivially from the algebraic rules for the addition of naturals.

@@

In the linear \picalc{}, the capabilities of the variables in context need to be modified every time input or output occurs.
However, the types of the variables in context need not.
As neither capabilities nor their indices depend on types, we choose to keep these two kinds of information separate (Figure \ref{contexts}): a length-indexed context of types $\PreCtx_n$ on one side; a length-indexed context of indices $\Idxs_n$ on the other, with a context of capabilities $\Ctx_{idxs}$ on top.

\begin{figure}[h]
  \begin{mathpar}
    \datatype
        {n : \N}
        {\PreCtx_n : \Set}

@@

        \inferrule
            {\gamma : \PreCtx_n \\ t : \Type}
            {\gamma \comma t : \PreCtx_{\suc n}}
  \end{mathpar}

  \begin{mathpar}
    \datatype
        {n : \N}
        {\Idxs_n : \Set}

@@

                \inferrule
                    {\Gamma : \Ctx_{idxs} \\ x : \Carrier_{idx}}
                    {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed typing contexts; length-indexed index context; index-indexed capability contexts.}
  \label{contexts}
\end{figure}

We represent base types $\base{n}$ using a natural number $n$ that can be reified into a type by the host language.
Using natural numbers instead of type universes is convenient: there is no need to litter type types with universe polymorphism, and it provides an additional proof of inhabitance, needed for the $\new \PO_{\suc n} \eqeq \PO_n$ rule in subject congruence.

  \inferrule
  {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Carrier_{idx}}}
  {\channel{t}{x} : \Type}
\end{mathpar}
\caption{Types for a resource-aware \picalc{}.}
\label{types}
\end{figure}

@@

@@

  \datatype{
    \gamma : \PreCtx_n \\
    \stacked{
      idxs : \Idxs_n \\\\
      \Gamma : \Ctx_{idxs}} \\
    i : \Var_n \\
    t : \Type \\
    \stacked{
      idx : \Idx \\\\
      x : \Carrier_{idx}} \\
    \Delta : \Ctx_{idxs}}
  {\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta} : \Set}

  \inferrule
      {\gamma : \PreCtx_n \\
        \stacked{idxs : \Idxs_n \\\\ \Gamma : \Ctx_{idxs}}\\
        t : \Type \\
        \stacked{idx : \Idx \\\\ x \; y \; z : \Carrier_{idx}}\\
        \op{x}{y}{z}}
  {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

  \inferrule
      {\gamma : \PreCtx_n \\
        \stacked{idxs : \Idxs_n \\\\ \Gamma : \Ctx_{idxs}}\\
        t \; t' \; : Type \\
        \stacked{idx : \Idx \\\\ x : \Carrier_{idx}} \\
        \stacked{idx' : \Idx \\\\ x' : \Carrier_{idx'}} \\
        \Delta : \Ctx_{idxs} \\\\
    loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
  {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
\end{mathpar}
\end{adjustwidth}
\caption{This is...}
\end{figure}

\todo{comment on scope restriction rule and capabilities}

@@

  \inferrule
  {t : \Type \\ x : \Carrier_{idx} \\ y : \Carrier_{idx'} \\\\
   cont : \types{\gamma \comma \channel{t}{x}}{\Gamma \comma y}{P}{\Delta \comma \lz}}
  {\constr{chan} \; t \; x \; y \; cont : \types{\gamma}{\Gamma}{\new P}{\Delta}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Progress}\label{progress}

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\Balanced                &:{} &                      &        & C \to \Set \\
  &\field{Balanced-}\lz     &:{} &                      &        & \Balanced \; \lz \\
  &\field{Balanced-}\lio    &:{} &                      &        & \Balanced \; \lio \\
  &\field{Balanced-\cdot^l} &:{} &\forall \{x y z\}     & \to \; & \Balanced \; x \to \Balanced \; z \to \op{x}{y}{z} \to \Balanced \; y \\
  &\field{Balanced?}        &:{} &\forall x             &        & \type{DEC} \; (\field{Balanced} \; x)
\end{aligned}
\end{equation}
\caption{}
\label{progress-capabilities}
\end{figure}

\todo{scope extension rule}

>>>>

% Fields
\newcommand{\op}[3]{#1 \; \field{\coloneqq} \; #2 \; \field{\cdot} \; #3}
\newcommand{\zero}{\field{0\cdot}}
\newcommand{\one}{\field{1\cdot}}
\newcommand{\li}{\field{\ell_i}}
\newcommand{\lo}{\field{\ell_o}}
\newcommand{\lz}{\field{\ell_{\o}}}
\newcommand{\lio}{\field{\ell_{\#}}}

@@

In the linear \picalc{} each channel has an input and an output \emph{capability}, and each capability has a given \emph{multiplicity} of 0 or 1.
In this paper we generalize over linearity and provide an algebra on multiplicities that is capable of encoding shared and gradual types as well.
The algebra has a single binary operator $(\cdot)$ that crushes multiplicities together.
It must satisfy the following constraints:
\begin{description}

\item [Partiality]
  Not all multiplicities are necessarily compatible: in the linear \picalc{} $1$ cannot be split into $1$ and $1$.
\item [Functionality]
  Given two multiplicities $y$ and $z$, it must be computable whether a third multiplicity $x$ exists such that $\op{x}{y}{z}$.
  If such an $x$ exists, it must be unique.

\item [Injectivity]
  To prove framing (\S \ref{framing}) we need to be able to \emph{subtract} multiplicities.
  For that end, we rely on $x - z \equiv y$ iff $\op{x}{y}{z}$.
  As a consequence, we must constraint $y$ to be uniquely determined as well.

@@

\item [Distinguished elements]
  There must be a neutral element $0$ absorbed on both sides.
  There must be a special element $1$ to record usage.
\end{description}

Together, these constraints give raise to a \emph{partial commutative monoid}.
In Figure \ref{fig:multiplicities} we present an algebra $\Quantifier_C$ on the carrier set $C$ in the form of a terniary relation such that it forms a partially commutative monoid.

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\zero                  &:{} &                      &        & C \\
  &\one                   &:{} &                      &        & C \\
  &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
  &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
  &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  &\field{\cdot-id^l}     &:{} &\forall x             & \to \; & \op{x}{\zero}{x} \\
  &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
  &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
\end{aligned}
\end{equation}
\caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
\label{fig:multiplicities}
\end{figure}

To allow for several multiplicity algebras to coexist on the same type system, we index the carriers of the algebras by an index $\Idx$ (see Figure \ref{fig:indexed-multiplicities}).
Each index $i$ has a corresponding carrier type $\field{CARRIER}_i$ such that $\field{CARRIER}_i$ satisfies the algebra $\Quantifier_{\field{CARRIER}_i}$.
We require the indexing type to be inhabited -- necessary to prove type preservation on $\new \PO_{\suc n} \eqeq \PO_n$.

\begin{figure}[h]
\begin{equation}
\begin{split}
  &\field{IDX}          &:{} &\Set \\
  &\field{\exists IDX}  &:{} &\field{IDX} \\
  &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
  &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
\end{split}
\end{equation}
\caption{Indexed set of partial commutative monoids.}
\label{fig:indexed-multiplicities}
\end{figure}

@@

\todo{Add typing derivations for example processes}

@@

The carrier $C$ is implemented as a type $\type{One}$ with a single trivial constructor $\constr{\omega}$.
Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
All other laws are trivially satisfied.

@@

The carrier $C$ is implemented as a type $\type{Two}$ with the trivial elements $\constr{zero}$ and $\constr{one}$.
The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{\one}$ -- and is uninhabited in every other case.
All other rules follow trivially.

@@

The carrier $C$ is implemented as the type of natural numbers $\N$.
The element $\zero$ corresponds 0, and $\one$ to 1.
The partial monoid $\op{x}{y}{z}$ is defined exactly when $x \equiv y + z$.
All other rules follow trivially from the algebraic rules for the addition of naturals.

@@

In a resource-aware \picalc{} type multiplicities change every time input and output occurs; however, the types themselves remain constant.
If multiplicities need not to depend on types, keeping types and multiplicities in independent contexts allows for a more modular approach to the algebra on contexts of multiplicities.
In this paper we choose that approach, which entails accepting that base non-channel types have dedicated input and output multiplicities too -- we will take care of impeding that any actual input or output can occur through them in the typing rules.

Figure \ref{fig:prectx} represents base types $\base{n}$ using a natural number $n$ that can be reified into a type by the host language.
Using natural numbers instead of type universes is convenient: there is no need to litter type types with universe polymorphism, and it provides an additional proof of inhabitance, needed for the $\new \PO_{\suc n} \eqeq \PO_n$ rule in subject congruence.

Types are then listed in a length-indexed \emph{pre-context} $\PreCtx_n$.

  \inferrule
  {t : \Type \\ \stacked{idx : \Idx \\\\ x : \Carrier_{idx}}}
  {\channel{t}{x} : \Type}
\end{mathpar}

\begin{mathpar}
  \datatype
      {n : \N}
      {\PreCtx_n : \Set}

@@

      \inferrule
          {\gamma : \PreCtx_n \\ t : \Type}
          {\gamma \comma t : \PreCtx_{\suc n}}
\end{mathpar}
\caption{Types and length-indexed typing contexts.}
\label{fig:prectx}
\end{figure}

To allow different multiplicity algebras to coexist we first define a length-indexed context of carrier indices $\Idxs_n$ (see \ref{fig:ctx}).
On top of it, we define a context of multiplicities $\Ctx_{idxs}$.
We lift the algebra on multiplicities to multiplicity contexts by constraining the algebra on multiplicity contexts to act on  contexts defined on the same underlying context of carrier indices.
We use two multiplicities over a common index to represent multiplicities for both the input and the output capabilities ($\Carrier_{idx}^2$).

\begin{figure}[h]
\begin{mathpar}
  \datatype
      {n : \N}
      {\Idxs_n : \Set}

@@

  \inferrule
      {\Gamma : \Ctx_{idxs} \\ x : \Carrier_{idx} ^2}
      {\Gamma \comma x : \Ctx_{idxs \comma idx}}
\end{mathpar}
\caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
\label{fig:ctx}
\end{figure}

@@

\todo{lift op to $^2$}

@@

  \datatype{
    \gamma : \PreCtx_n \\
    \stacked{
      idxs : \Idxs_n \\\\
      \Gamma : \Ctx_{idxs}} \\
    i : \Var_n \\
    t : \Type \\
    \stacked{
      idx : \Idx \\\\
      x : \Carrier_{idx}^2} \\
    \Delta : \Ctx_{idxs}}
  {\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta} : \Set}

  \inferrule
      {\gamma : \PreCtx_n \\
        \stacked{idxs : \Idxs_n \\\\ \Gamma : \Ctx_{idxs}}\\
        t : \Type \\
        \stacked{idx : \Idx \\\\ x \; y \; z : \Carrier_{idx}^2}\\
        \op{x}{y}{z}}
  {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

  \inferrule
      {\gamma : \PreCtx_n \\
        \stacked{idxs : \Idxs_n \\\\ \Gamma : \Ctx_{idxs}}\\
        t \; t' \; : Type \\
        \stacked{idx : \Idx \\\\ x : \Carrier_{idx}^2} \\
        \stacked{idx' : \Idx \\\\ x' : \Carrier_{idx'}^2} \\
        \Delta : \Ctx_{idxs} \\\\
    loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
  {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
\end{mathpar}
\end{adjustwidth}
\caption{This is...}
\end{figure}

\todo{comment on balancing in chan rule}

@@

  \inferrule
  {t : \Type \\ x : \Carrier_{idx}^2 \\ y : \Carrier_{idx'} \\\\
   cont : \types{\gamma \comma \channel{t}{x}}{\Gamma \comma (y \comma y) }{P}{\Delta \comma \lz}}
  {\constr{chan} \; t \; x \; y \; cont : \types{\gamma}{\Gamma}{\new P}{\Delta}}

@@
